---
format: md
---

# BCB784-proj


Data analysis is distributed amongst other R scripts, some of which use the 
[box](https://github.com/klmr/box) package syntax to make R scripts into 
reproducible modules (similar to python)

### Data Availability

Small files are stored in the `/data` directory.

Data files that are too large for github may be temporarily found via [google drive](https://drive.google.com/file/d/1kA661pwAmi0wLTlGC3YhXbE5J_nwMui2/view?usp=drive_link)

Source data may be found from the gtex portal with the 
[long read](https://gtexportal.org/home/downloads/adult-gtex/long_read_data) and
[short read](https://gtexportal.org/home/downloads/adult-gtex/bulk_tissue_expression)
data.

## loading data

```{r}
box::use(
  # ensures packages are loaded and attached
  # if `plyxp` is not be available for the user, in which
  # case they should either upgrade their `BiocManager`
  # version to 3.20, or use `remotes::install_github("jtlandis/plyxp")`
  analysis/dependencies[...],
  # makes various datasets available
  analysis/datasets
)

# long read data
datasets$se_long
# short read data
datasets$se_short
```

### datasets annotations

Please see [`analysis/datasets.R`] as most values exported are annotated there.

### Subsetting short read Samples

Please see [`data/filter_short_read_bulk_RNAseq.R`] for source code on subsetting
large rna-seq samples.

## Data Summary

The GTEX short read data contains many different versions, the two most recent 
being version 8 and version 10. Given that the 
[associated paper](https://www.nature.com/articles/s41586-022-05035-y)
stated they used Gencode Version 26, we opted to use GTEX version 8. The long
read data only has one GTEX version available, version 9, which is not annotated
as to which Gencode version was used.

## Figure Reproduction

```{r}
box::use(f1c = analysis/fig1c,
         ggplot2[...])

f1c$plot_figc(f1c$plot_data_transcripts) +
  ggtitle("Transcripts")
f1c$plot_figc(f1c$plot_data_genes) +
  ggtitle("Genes")
```








